---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE
)
```

# btrank

**btrank** is an R package for estimating latent ability scores from paired 
comparison data using the **Bradley-Terry model**. It is designed to be 
domain-agnostic — applicable to sports rankings, preference experiments, 
academic evaluations, or any setting where items are compared pairwise 
across multiple periods.

## Key features

- Flexible input: accepts any data frame with item, period, and score columns
- Temporal weighting via exponential decay (Ley et al., 2019)
- Win probability matrix between all pairs of items
- Modular pipeline and a convenient one-step wrapper
- Full roxygen2 documentation and a worked vignette

## Installation

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("alvaromkm/btrank")
```

## Quick example

```{r example}
library(btrank)

results <- data.frame(
  item   = c("A", "B", "C", "A", "B", "C"),
  period = c(2022, 2022, 2022, 2023, 2023, 2023),
  score  = c(80, 65, 50, 75, 85, 60)
)

# Full pipeline in one call
bt_rank_all(results, score_col = "score")

# With temporal weighting — recent periods count more
bt_rank_all(results, score_col = "score", half_life = 1)
```

## Pipeline overview

| Function | Description |
|---|---|
| `bt_win_matrix()` | Build pairwise win count matrix |
| `bt_fit()` | Fit Bradley-Terry model |
| `bt_rank()` | Extract ranked data frame |
| `bt_prob_matrix()` | Compute win probability matrix |
| `bt_weights()` | Generate exponential decay weights |
| `bt_rank_all()` | Full pipeline in one call |

## Vignette

A full worked example with methodological background is available at:

```{r vignette, eval = FALSE}
browseVignettes("btrank")
```

## References

- Bradley, R. A., & Terry, M. E. (1952). Rank analysis of incomplete block 
designs. *Biometrika*, 39(3/4), 324–345.
- Ley, C., et al. (2019). Ranking soccer teams on the basis of their current 
strength. *Statistical Modelling*, 19(1), 55–73.
- Alvo, M., & Yu, P. L. H. (2014). *Statistical Methods for Ranking Data*. 
Springer. https://doi.org/10.1007/978-1-4939-1471-5
- Yu, P. L. H., Gu, J., & Xu, H. (2019). Analysis of ranking data. 
*WIREs Computational Statistics*. https://doi.org/10.1002/wics.1483

## License

MIT © Alvaro Meca Mondejar
