---
title: "Introduction to btrank"
author: "Álvaro Meca Mondéjar, Miguel Ángel Montero Alonso, Juan de Dios Luna del Castillo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to btrank}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

## Overview

**btrank** is built on the Bradley-Terry model (Bradley & Terry, 1952), a probabilistic model for paired comparisons. Each item $i$ is assigned a latent ability $\lambda_i$, a parameter that is not observed directly but
inferred from the outcomes of comparisons between items. The probability that $i$ beats $j$ depends only on the difference between their abilities:

$$P(i \text{ beats } j) = \frac{e^{\lambda_i}}{e^{\lambda_i} + e^{\lambda_j}}$$

The package supplies a small, composable set of functions around this
model: building pairwise comparisons from repeated observations
(`bt_win_matrix()`), fitting the model (`bt_fit()`), extracting and
comparing results (`bt_rank()`, `bt_prob_matrix()`), quantifying
uncertainty (`summary.btfit()`, `confint.btfit()`,
`bt_significance_matrix()`), and visualising a fit (`plot.btfit()`).

This vignette's purpose is to document what each function does, what its
arguments control, and what it returns — using 31 seasons of English
Premier League results as a running example, because its outcomes are
independently checkable. The football results themselves are illustration,
not the subject.

## Installation

```{r install, eval = FALSE}
remotes::install_github("alvaromkm/btrank")
```

```{r load}
library(btrank)
```

## Input data format

Every function in **btrank** starts from a data frame with at most one row
per item per period:

| Column | Description |
|--------|-------------|
| `item` | Item identifier |
| `period` | Comparison block: the unit within which scores are comparable |
| `score` | Numeric metric used to rank items within each period |

A "period" need not be a unit of time. It is any block within which scores
can be compared: a season (items are teams, score is points), a single match
(two rows, score is goals), a consumer (items are products, score is a
rating) or a judge (items are wines, score is a mark). Scores are never
compared across periods, so an item may appear only once per period; for
ratings from several consumers, the consumer, not the month, is the period.

Column names are configurable via `item_col`, `period_col`, `score_col`
wherever they appear. By default higher `score` is better;
`higher_is_better = FALSE` reverses this (e.g. for times or error counts).

## The EPL dataset

```{r load-data}
data(epl_scores)
dim(epl_scores)
head(epl_scores, 4)
```

`epl_scores` has one row per team per season (1995/96-2025/26): `points`
(3/1/0) and `goal_diff`, already in `item`/`period`/`score` form.

## Data provenance and reproducibility

Match results come from
[football-data.co.uk](https://www.football-data.co.uk/englandm.php)
(division "E0"). Raw match-level files are not redistributed with this
package — only the aggregated `epl_scores` table is shipped. The
aggregation itself is the internal, unit-tested function
`.matches_to_scores()`; the exact steps to regenerate `epl_scores` from
source are in `data-raw/prepare_epl_scores.R`.

## Building the win matrix: `bt_win_matrix()`

`bt_win_matrix()` converts item/period/score data into a matrix of
pairwise win counts — the input `bt_fit()` needs. Within each period, it
ranks items by `score` and records a win for the higher-ranked item over
every lower-ranked item; exact ties split 0.5/0.5. Its arguments:

- `item_col`, `period_col`, `score_col` — column names (defaults
  `"item"`, `"period"`, `"score"`).
- `higher_is_better` — direction of `score`. With `TRUE` (the default) the
  item with the larger score in a period wins the comparison; with `FALSE`
  the smaller score wins, as is appropriate for lap times or error counts.
  Items with exactly equal scores in a period split the comparison 0.5/0.5.
- `weights` — optional named vector (or list) with one weight per period,
  keyed by the period value. Every comparison generated within period $t$
  is counted with weight $w_t$ instead of 1, so that some periods
  contribute more evidence than others. The weights are typically produced
  by `bt_weights()` (see below); the resulting win counts are then no
  longer integers.
- `absent` — how to treat an item that does not appear in a period at all
  (for instance, a team that was not in the league that season). With
  `"ignore"` (the default) such an item simply takes no part in that
  period. With `"penalize"`, every item present in the period is
  additionally credited `absent_penalty` wins (default 1) over each item
  absent from it; no comparison is generated between two absent items.
  When `weights` are supplied, this credit is scaled by the weight of the
  period. This vignette uses `"penalize"` throughout, except in the section
  devoted to comparing both options. It only makes sense when every period
  is a full participation unit, such as a league season.
- `cluster_col` — optional column identifying the independent units used
  for standard errors (see *Quantifying uncertainty*). By default each
  period is its own cluster; a different column is needed when several
  periods share a source of dependence, e.g. one period per pairwise
  judgement but many judgements per respondent.
- `time_col` — optional column whose values are matched to the names of
  `weights`, for when periods are not time units (e.g. one period per match,
  weighted by season).

```{r win-matrix}
win_mat_points <- bt_win_matrix(epl_scores, score_col = "points", absent = "penalize")
dim(win_mat_points)
win_mat_points["Man United", "Arsenal"]
win_mat_points["Arsenal", "Man United"]
```

The matrix is not symmetric: entry `[i, j]` is the number of times `i`
outranked `j`. It also keeps, as the attribute `"comparisons"`, one row per
pairwise comparison with the period and cluster it came from; the win
counts alone cannot tell which comparisons share a period, and the
standard errors below need that information.

## Fitting the model: `bt_fit()`

`bt_fit()` takes a win-count matrix and fits the model via
`BradleyTerry2::BTm()`, in two internal passes: an arbitrary reference is
used first to identify the weakest item, then the model is re-fit with
that item as reference, so every reported ability is $\geq 0$.

```{r fit-points}
fit_points <- bt_fit(win_mat_points)
print(fit_points)
```

The returned `btfit` object is a list: `abilities` (named vector),
`reference` (the anchor item), `items`, `model` (the underlying `BTm`
fit, for advanced use), `separation_items` — items that never won or
never lost a single comparison, flagged automatically (see the
Uncertainty section below for why this matters) — and `comparisons`, the
comparison-level record taken from the win matrix. Here `separation_items`
is empty: even with `absent = "penalize"` applied over the full 31-year,
49-team universe, every item retains at least one real win.

## Extracting a ranking: `bt_rank()`

```{r rank-points}
bt_rank(fit_points)[1:6, ]
```

`bt_rank()` returns `item`, `rank`, `ability` (raw $\lambda$), and
`ability_norm` — a $[0, 1]$ rescaling (weakest $\to$ 0, strongest $\to$
1). Because raw $\lambda$ is scale-dependent (it varies with the number
of items, periods, and separation), `ability_norm` is the right column
for comparing across different fits; `bt_rank(fit, top_n = k)` truncates
to the top `k`.

## Temporal weighting: `bt_weights()`

`bt_weights(periods, half_life)` returns exponential decay weights
(Ley et al., 2019):

$$w_t = \left(\tfrac{1}{2}\right)^{(T-t)/T_{1/2}}$$

with $T$ the most recent period; the most recent period always has
weight 1. The result is a named vector, passed directly to
`bt_win_matrix(weights = ...)`.

```{r weights}
periods  <- sort(unique(epl_scores$period))
w_recent <- bt_weights(periods, half_life = 5)

fit_recent <- bt_fit(bt_win_matrix(epl_scores, score_col = "points",
                                    weights = w_recent, absent = "penalize"))

merge(
  data.frame(item = names(fit_points$abilities), rank_flat = rank(-fit_points$abilities)),
  data.frame(item = names(fit_recent$abilities), rank_recent = rank(-fit_recent$abilities)),
  by = "item"
) |> subset(item %in% c("Man City", "Man United"))
```

A shorter `half_life` weights recent periods more heavily; here it moves
Manchester City from 6th to 1st and Manchester United from 1st to 4th —
the mechanism, not a claim about football, is the point.

## Multiple metrics

Any numeric column works as `score_col` — nothing else changes:

```{r fit-goal-diff}
fit_gd <- bt_fit(bt_win_matrix(epl_scores, score_col = "goal_diff", absent = "penalize"))

cor(rank(-fit_points$abilities)[names(fit_gd$abilities)],
    rank(-fit_gd$abilities), method = "spearman")
```

The two rankings correlate at $\approx 0.997$ here — near-total
agreement — yet even so, Arsenal ranks above Manchester United by goal
difference but below it by points: both computed by the identical
function, just a different column, capturing "who won more" versus "who
dominated more."

## Handling absence

`bt_win_matrix()`'s `absent` argument controls items missing from a
period entirely (promotion/relegation is the natural case): `"ignore"`
(default) contributes nothing for missing periods; `"penalize"` credits
every present item a win against each absent item, `absent_penalty`
times per missing period (default `1`, one full loss). The rest of this
vignette uses `"penalize"`; this section demonstrates why the choice
matters, comparing both explicitly on a bounded, recent window.

```{r absent}
recent <- epl_scores[epl_scores$period %in% 2021:2025, ]

fit_ignore   <- bt_fit(bt_win_matrix(recent, score_col = "points", absent = "ignore"))
fit_penalize <- bt_fit(bt_win_matrix(recent, score_col = "points",
                                      absent = "penalize", absent_penalty = 1))

rbind(
  ignore   = rank(-fit_ignore$abilities)["Sunderland"],
  penalize = rank(-fit_penalize$abilities)["Sunderland"]
)
```

Sunderland played one of these five seasons; under `"ignore"` its single
season alone puts it 4th, ahead of Manchester United; under
`"penalize"` it falls to 20th. Neither is a free default — `"ignore"`
can let one strong season outweigh years of absence; `"penalize"`
treats absence itself as informative, a modelling assumption that
should be stated explicitly, not left implicit.

## Quantifying uncertainty: `summary.btfit()`, `confint.btfit()`, `vcov.btfit()`

`bt_fit()`'s point estimates say nothing about precision. The usual
model-based standard errors of `BTm()` treat every comparison as an
independent binomial trial. That is not the case here: the
`r choose(20, 2)` comparisons generated from one 20-team season are all
functions of the same 20 scores. `vcov.btfit()` therefore returns, by
default, a **cluster-robust** (sandwich) covariance, which treats each
cluster (by default, each period) as the independent unit (Liang & Zeger,
1986). The fit is a composite likelihood whose point estimates are
unaffected; only their variance changes (Varin, Reid & Firth, 2011).
Temporal weights are handled as importance weights, not as frequencies.
`vcov(fit, type = "model")` gives the model-based estimate, for comparison
only.

`summary.btfit(object, level = 0.95)` builds a table of `item`,
`ability`, `se`, `ci_low`, `ci_high`, and `reliable`; `confint.btfit()`
returns just the interval, `stats`-style.

```{r summary}
s <- summary(fit_points, level = 0.95)
s_top <- s
s_top$table <- head(s$table, 8)   # print only the first rows
s_top
```

Three details matter when reading this table:

- **Quasi-standard errors.** Abilities are only defined relative to the
  reference item, so an ordinary standard error depends on that arbitrary
  choice. `se` is instead the quasi-standard error (Firth & de Menezes,
  2004), which does not depend on the reference and exists for every item,
  the reference included. The intervals are comparison intervals: they
  are for comparing items with each other, not for an absolute ability.
- **t reference.** Intervals use a t distribution with $G - 1$ degrees of
  freedom, $G$ being the number of clusters (here `r s$n_clusters`). The
  header also reports an *effective* number of clusters, which falls when
  temporal weighting concentrates the information in few periods; it is
  descriptive only.
- **Items in a single cluster.** The cluster-robust estimate cannot
  measure the uncertainty of an item whose comparisons all come from one
  cluster (a team present in a single season under `absent = "ignore"`):
  such items get `NA` and `reliable = FALSE`. With `absent = "penalize"`,
  as here, structural comparisons place every team in every season, so no
  item is affected.

A concrete question this answers: is Liverpool's ability significantly
different from Chelsea's? Their intervals —
[`r sprintf("%.2f, %.2f", s$table$ci_low[s$table$item == "Liverpool"], s$table$ci_high[s$table$item == "Liverpool"])`]
and
[`r sprintf("%.2f, %.2f", s$table$ci_low[s$table$item == "Chelsea"], s$table$ci_high[s$table$item == "Chelsea"])`]
— overlap substantially; overlapping intervals do not prove "no
difference" on their own (`bt_significance_matrix()` below tests this
properly), but they immediately show the gap is small relative to the
uncertainty, which the point estimates alone do not.

If any item is flagged in `separation_items`, `reliable` is `FALSE` for
every row: quasi-variances are fitted jointly to the whole covariance
matrix, so the diverging variance of a separated item can distort every
item's standard error.

## Pairwise win probabilities: `bt_prob_matrix()`

```{r prob-matrix}
pm <- bt_prob_matrix(fit_points)
pm["Man United", "Arsenal"]
pm["Man United", "Chelsea"]
```

`bt_prob_matrix(fit, order_by = "ability")` returns an $n \times n$
matrix where entry `[i, j]` is $P(i \text{ beats } j)$ under the fitted
model, diagonal `NA`. `order_by = "name"` sorts alphabetically instead.
Note this is a **model-implied** probability from the two abilities
alone, not an empirical win rate.

## Pairwise significance testing: `bt_significance_matrix()`

`bt_significance_matrix(fit, level = 0.95, method = "holm")` runs a Wald
test on every pairwise ability difference, using the full covariance from
`vcov.btfit()` (cluster-robust by default, with the same t reference as
`summary.btfit()`), so correlation between two items' estimates is
accounted for, not just their individual SEs. The variance of a difference
does not depend on the reference item, so these tests are exact with
respect to that choice. With $n$ items there are
$n(n-1)/2$ simultaneous comparisons, so p-values are adjusted by
default — `"holm"` (Holm, 1979), `"bonferroni"`, `"BH"`, or `"none"`
(provided only for comparison; unadjusted p-values understate the true
false-positive rate here). It returns `p`, `significant`, `method`,
`level`, `type`, `df`, and the same fit-wide `reliable` flag as
`summary.btfit()`.

**Is the "big six" actually six distinguishable teams?** Refitting on
just those six clubs (same two functions, restricted data — not a
special case):

```{r big-six}
big6 <- c("Man United", "Arsenal", "Liverpool", "Chelsea", "Man City", "Tottenham")
fit_big6 <- bt_fit(bt_win_matrix(epl_scores[epl_scores$item %in% big6, ],
                                  score_col = "points", absent = "penalize"))

sort(fit_big6$abilities, decreasing = TRUE)

sig6 <- bt_significance_matrix(fit_big6)
sig6
```

```{r big-six-count, include = FALSE}
p_rob <- sig6$p[upper.tri(sig6$p)]
sig6_model <- suppressWarnings(bt_significance_matrix(fit_big6, type = "model"))
p_mod <- sig6_model$p[upper.tri(sig6_model$p)]
```

Of the 15 pairs, `r sum(p_rob < 0.05)` are significant after Holm
correction: Tottenham against Manchester United, Arsenal, Chelsea and
Liverpool. Every other pair, including Tottenham against Manchester City,
is not. Tottenham is reliably behind four of the other five clubs; the
rest of the group cannot be separated.

The choice of standard error matters here. With the model-based errors,
which treat each season's 15 comparisons as independent,
`r sum(p_mod < 0.05)` pairs would be declared significant, including
Manchester City against Manchester United:

```{r big-six-model}
sum(p_mod < 0.05)
sum(p_rob < 0.05)
```

This is exactly the kind of claim a ranking table alone cannot support,
and the reason `bt_significance_matrix()` exists as a separate step from
`bt_rank()`.

## Visualizing model fit: `plot.btfit()`

`plot.btfit(x, level = 0.95, ...)` draws a caterpillar plot: one point
per item (ability), one horizontal segment per item (the comparison
interval from `summary.btfit()`), ordered weakest-to-strongest; items
flagged `reliable = FALSE` are drawn in grey rather than black. It is a thin wrapper around `graphics::plot()`
using exactly the table `summary.btfit()` computes — no separate
calculation.

```{r plot-btfit, eval = FALSE}
plot(fit_points)
```


## Relationship to BradleyTerry2

`bt_fit()` does not implement its own estimator: it calls
`BradleyTerry2::BTm()` internally. To check this, the chunk below fits the
same win matrix directly with `BTm()` and compares the two sets of abilities.

```{r vs-bradleyterry2}
items <- rownames(win_mat_points)
pairs <- combn(items, 2)
df_manual <- data.frame(
  item1 = factor(pairs[1, ], levels = items), item2 = factor(pairs[2, ], levels = items),
  wins1 = apply(pairs, 2, function(p) win_mat_points[p[1], p[2]]),
  wins2 = apply(pairs, 2, function(p) win_mat_points[p[2], p[1]])
)
fit_manual <- BradleyTerry2::BTm(outcome = cbind(wins1, wins2), player1 = item1, player2 = item2, data = df_manual)

ability_manual <- BradleyTerry2::BTabilities(fit_manual)[, 1]
ability_btrank <- fit_points$abilities[names(ability_manual)]
common_ref <- names(ability_manual)[1]
max(abs((ability_manual - ability_manual[common_ref]) - (ability_btrank - ability_btrank[common_ref])))
```

The largest difference is of the order of $10^{-15}$, that is, zero up to
rounding error. One detail is needed for the comparison: abilities are only
defined relative to a reference item whose ability is set to 0. `BTm()` uses
the first item, while `bt_fit()` uses the weakest. Both fits are therefore
shifted to a common reference before comparing; without that step the two
sets of values differ by a constant, even though they describe the same model.

What **btrank** adds is everything around the estimator:

- building the comparisons from repeated scores (`bt_win_matrix()`);
- temporal weighting (`bt_weights()`);
- handling of items absent from a period (`absent`);
- standard errors that account for the dependence between comparisons
  from the same period, reported as reference-free quasi-standard errors
  (`vcov.btfit()`, `summary.btfit()`);
- warnings when standard errors are unreliable because of separation or
  single-cluster items, and pairwise significance tests with multiplicity
  correction (`bt_significance_matrix()`).


## Limitations

The following simplifications are known, and are the main directions for
further development.

- **Standard errors.** Cluster-robust standard errors assume that clusters
  are independent of each other and rely on having many of them; with few
  clusters, or with strong temporal decay, they can be too small. Items
  present in a single cluster get no standard error at all. Their
  finite-sample coverage has not yet been assessed by simulation.
- **Temporal weighting.** `half_life` is chosen by the user, not estimated,
  and the resulting rankings depend on it (see the section on temporal
  weighting). Results should be checked over a range of values; selecting it
  from the data, for instance by predictive validation, is left for future
  work.
- **Home advantage.** The model has no home-advantage parameter. With
  season-level scores its effect largely cancels out, since every team plays
  the same number of home and away matches, but it would matter for
  match-level data (one period per match).
- **Separation.** Items that never win or never lose are flagged, but their
  estimates are not corrected. A bias-reduced estimator such as Firth's
  (1993) is a natural next step.
- **Aggregation.** Only the order of scores within each period is used, not
  the individual matches or the size of the gaps between scores: winning a
  league by twenty points counts the same as winning it by one. This is what
  keeps weighting across periods well defined, but it discards information.
  Match-level data can be analysed by using each match as a period, at the
  cost of the season-level summary.

## References

Bradley, R. A., & Terry, M. E. (1952). Rank analysis of incomplete block
designs: I. The method of paired comparisons. *Biometrika*, 39(3/4), 324–345.

Turner, H., & Firth, D. (2012). Bradley-Terry models in R: The BradleyTerry2
package. *Journal of Statistical Software*, 48(9), 1–21.
https://doi.org/10.18637/jss.v048.i09

Ley, C., Van de Wiele, T., & Van Eetvelde, H. (2019). Ranking soccer teams
on the basis of their current strength: A comparison of maximum likelihood
approaches. *Statistical Modelling*, 19(1), 55–77.

Holm, S. (1979). A simple sequentially rejective multiple test procedure.
*Scandinavian Journal of Statistics*, 6(2), 65–70.

Liu, H. H., Battaglia, J., & Wu, T. T. (2025). Impact of COVID-19 on
English Football Premier League: Analyzing rankings and home advantage
using extended Bradley-Terry models. *PLOS One*, 20(10), e0332627.
https://doi.org/10.1371/journal.pone.0332627

Firth, D., & de Menezes, R. X. (2004). Quasi-variances. *Biometrika*,
91(1), 65–80.

Liang, K.-Y., & Zeger, S. L. (1986). Longitudinal data analysis using
generalized linear models. *Biometrika*, 73(1), 13–22.

Varin, C., Reid, N., & Firth, D. (2011). An overview of composite
likelihood methods. *Statistica Sinica*, 21(1), 5–42.

Firth, D. (1993). Bias reduction of maximum likelihood estimates.
*Biometrika*, 80(1), 27–38.

football-data.co.uk. English Premier League results, 1995/96–2025/26.
https://www.football-data.co.uk/englandm.php