---
title: "btrank: Bradley-Terry Rankings for Paired Comparison Data"
author: "Álvaro Meca Mondéjar, Miguel Ángel Montero Alonso, Juan de Dios Luna del Castillo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{btrank: Bradley-Terry Rankings for Paired Comparison Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

## Overview

**btrank** estimates latent ability scores from paired comparison data using
the Bradley-Terry model (Bradley & Terry, 1952). Each item $i$ has a latent
ability $\lambda_i$; for two items $i, j$:

$$P(i \text{ beats } j) = \frac{e^{\lambda_i}}{e^{\lambda_i} + e^{\lambda_j}}$$

The package supplies a small, composable set of functions around this
model: building pairwise comparisons from repeated observations
(`bt_win_matrix()`), fitting the model (`bt_fit()`), extracting and
comparing results (`bt_rank()`, `bt_prob_matrix()`), quantifying
uncertainty (`summary.btfit()`, `confint.btfit()`,
`bt_significance_matrix()`), and visualising a fit (`plot.btfit()`).

This vignette's purpose is to document what each function does, what its
arguments control, and what it returns — using 31 seasons of English
Premier League results as a running example, because its outcomes are
independently checkable. The football results themselves are illustration,
not the subject.

## Installation

```{r install, eval = FALSE}
remotes::install_github("alvaromkm/btrank")
```

```{r load}
library(btrank)
```

## Input data format

Every function in **btrank** starts from a data frame with one row per
item per period:

| Column | Description |
|--------|-------------|
| `item` | Item identifier |
| `period` | Period identifier (season, year, round, ...) |
| `score` | Numeric metric used to rank items within each period |

Column names are configurable via `item_col`, `period_col`, `score_col`
wherever they appear. By default higher `score` is better;
`higher_is_better = FALSE` reverses this (e.g. for times or error counts).

## The EPL dataset

```{r load-data}
data(epl_scores)
dim(epl_scores)
head(epl_scores, 4)
```

`epl_scores` has one row per team per season (1995/96-2025/26): `points`
(3/1/0) and `goal_diff`, already in `item`/`period`/`score` form.

## Data provenance and reproducibility

Match results come from
[football-data.co.uk](https://www.football-data.co.uk/englandm.php)
(division "E0"). Raw match-level files are not redistributed with this
package — only the aggregated `epl_scores` table is shipped. The
aggregation itself is the internal, unit-tested function
`.matches_to_scores()`; the exact steps to regenerate `epl_scores` from
source are in `data-raw/prepare_epl_scores.R`.

## Building the win matrix: `bt_win_matrix()`

`bt_win_matrix()` converts item/period/score data into a matrix of
pairwise win counts — the input `bt_fit()` needs. Within each period, it
ranks items by `score` and records a win for the higher-ranked item over
every lower-ranked item; exact ties split 0.5/0.5. Its arguments:

- `item_col`, `period_col`, `score_col` — column names (defaults
  `"item"`, `"period"`, `"score"`).
- `higher_is_better` — direction of `score`.
- `weights` — optional named vector/list, one weight per period, applied
  to every comparison in that period (see `bt_weights()` below).
- `absent` — how to treat an item missing from a period entirely (see
  below). Used here as `"penalize"` throughout, except where `"ignore"`
  is the explicit subject of a section.

```{r win-matrix}
win_mat_points <- bt_win_matrix(epl_scores, score_col = "points", absent = "penalize")
dim(win_mat_points)
win_mat_points["Man United", "Arsenal"]
win_mat_points["Arsenal", "Man United"]
```

The matrix is not symmetric: entry `[i, j]` is the number of times `i`
outranked `j`.

## Fitting the model: `bt_fit()`

`bt_fit()` takes a win-count matrix and fits the model via
`BradleyTerry2::BTm()`, in two internal passes: an arbitrary reference is
used first to identify the weakest item, then the model is re-fit with
that item as reference, so every reported ability is $\geq 0$.

```{r fit-points}
fit_points <- bt_fit(win_mat_points)
print(fit_points)
```

The returned `btfit` object is a list: `abilities` (named vector),
`reference` (the anchor item), `items`, `model` (the underlying `BTm`
fit, for advanced use), and `separation_items` — items that never won or
never lost a single comparison, flagged automatically (see the
Uncertainty section below for why this matters). Here `separation_items`
is empty: even with `absent = "penalize"` applied over the full 31-year,
49-team universe, every item retains at least one real win.

## Extracting a ranking: `bt_rank()`

```{r rank-points}
bt_rank(fit_points)[1:6, ]
```

`bt_rank()` returns `item`, `rank`, `ability` (raw $\lambda$), and
`ability_norm` — a $[0, 1]$ rescaling (weakest $\to$ 0, strongest $\to$
1). Because raw $\lambda$ is scale-dependent (it varies with the number
of items, periods, and separation), `ability_norm` is the right column
for comparing across different fits; `bt_rank(fit, top_n = k)` truncates
to the top `k`.

## Temporal weighting: `bt_weights()`

`bt_weights(periods, half_life)` returns exponential decay weights
(Ley et al., 2019):

$$w_t = \left(\tfrac{1}{2}\right)^{(T-t)/T_{1/2}}$$

with $T$ the most recent period; the most recent period always has
weight 1. The result is a named vector, passed directly to
`bt_win_matrix(weights = ...)`.

```{r weights}
periods  <- sort(unique(epl_scores$period))
w_recent <- bt_weights(periods, half_life = 5)

fit_recent <- bt_fit(bt_win_matrix(epl_scores, score_col = "points",
                                    weights = w_recent, absent = "penalize"))

merge(
  data.frame(item = names(fit_points$abilities), rank_flat = rank(-fit_points$abilities)),
  data.frame(item = names(fit_recent$abilities), rank_recent = rank(-fit_recent$abilities)),
  by = "item"
) |> subset(item %in% c("Man City", "Man United"))
```

A shorter `half_life` weights recent periods more heavily; here it moves
Manchester City from 6th to 1st and Manchester United from 1st to 4th —
the mechanism, not a claim about football, is the point.

## Multiple metrics

Any numeric column works as `score_col` — nothing else changes:

```{r fit-goal-diff}
fit_gd <- bt_fit(bt_win_matrix(epl_scores, score_col = "goal_diff", absent = "penalize"))

cor(rank(-fit_points$abilities)[names(fit_gd$abilities)],
    rank(-fit_gd$abilities), method = "spearman")
```

The two rankings correlate at $\approx 0.997$ here — near-total
agreement — yet even so, Arsenal ranks above Manchester United by goal
difference but below it by points: both computed by the identical
function, just a different column, capturing "who won more" versus "who
dominated more."

## Handling absence

`bt_win_matrix()`'s `absent` argument controls items missing from a
period entirely (promotion/relegation is the natural case): `"ignore"`
(default) contributes nothing for missing periods; `"penalize"` credits
every present item a win against each absent item, `absent_penalty`
times per missing period (default `1`, one full loss). The rest of this
vignette uses `"penalize"`; this section demonstrates why the choice
matters, comparing both explicitly on a bounded, recent window.

```{r absent}
recent <- epl_scores[epl_scores$period %in% 2021:2025, ]

fit_ignore   <- bt_fit(bt_win_matrix(recent, score_col = "points", absent = "ignore"))
fit_penalize <- bt_fit(bt_win_matrix(recent, score_col = "points",
                                      absent = "penalize", absent_penalty = 1))

rbind(
  ignore   = rank(-fit_ignore$abilities)["Sunderland"],
  penalize = rank(-fit_penalize$abilities)["Sunderland"]
)
```

Sunderland played one of these five seasons; under `"ignore"` its single
season alone puts it 4th, ahead of Manchester United; under
`"penalize"` it falls to 20th. Neither is a free default — `"ignore"`
can let one strong season outweigh years of absence; `"penalize"`
treats absence itself as informative, a modelling assumption that
should be stated explicitly, not left implicit.

## Quantifying uncertainty: `summary.btfit()`, `confint.btfit()`, `vcov.btfit()`

`bt_fit()`'s point estimates say nothing about precision. `vcov.btfit()`
extracts the variance-covariance matrix of the fitted abilities (from
`BTm()`'s underlying `glm()`); `summary.btfit(object, level = 0.95)`
builds a table of `item`, `ability`, `se`, `ci_low`, `ci_high`, and
`reliable`; `confint.btfit()` returns just the interval, `stats`-style.

```{r summary}
s <- summary(fit_points, level = 0.95)
s$table[s$table$item %in% c("Man United", "Arsenal", "Liverpool", "Chelsea"), ]
```

A concrete question this answers: is Liverpool's ability significantly
different from Chelsea's? Their Wald intervals — [6.31, 7.25] and
[6.21, 7.14] — overlap substantially; overlapping intervals do not
prove "no difference" on their own (`bt_significance_matrix()` below
tests this properly), but they immediately show the gap is small
relative to the uncertainty, which the point estimates alone do not.

`reliable` applies to the **whole fit**, not per item: `bt_fit()` fixes
the *weakest* item as reference, and a never-winning item is almost
always that weakest item, so a broken standard error for the reference
propagates identically to every other item's SE (each is expressed
relative to it). If any item is flagged in `separation_items`,
`reliable` is `FALSE` for every row — reported, not silently omitted.

## Pairwise win probabilities: `bt_prob_matrix()`

```{r prob-matrix}
pm <- bt_prob_matrix(fit_points)
pm["Man United", "Arsenal"]
pm["Man United", "Chelsea"]
```

`bt_prob_matrix(fit, order_by = "ability")` returns an $n \times n$
matrix where entry `[i, j]` is $P(i \text{ beats } j)$ under the fitted
model, diagonal `NA`. `order_by = "name"` sorts alphabetically instead.
Note this is a **model-implied** probability from the two abilities
alone, not an empirical win rate.

## Pairwise significance testing: `bt_significance_matrix()`

`bt_significance_matrix(fit, level = 0.95, method = "holm")` runs a Wald
test on every pairwise ability difference, using `vcov.btfit()` for the
joint covariance (so correlation between two items' estimates is
accounted for, not just their individual SEs). With $n$ items there are
$n(n-1)/2$ simultaneous comparisons, so p-values are adjusted by
default — `"holm"` (Holm, 1979), `"bonferroni"`, `"BH"`, or `"none"`
(provided only for comparison; unadjusted p-values understate the true
false-positive rate here). It returns `p`, `significant`, `method`,
`level`, and the same fit-wide `reliable` flag as `summary.btfit()`.

**Is the "big six" actually six distinguishable teams?** Refitting on
just those six clubs (same two functions, restricted data — not a
special case):

```{r big-six}
big6 <- c("Man United", "Arsenal", "Liverpool", "Chelsea", "Man City", "Tottenham")
fit_big6 <- bt_fit(bt_win_matrix(epl_scores[epl_scores$item %in% big6, ],
                                  score_col = "points", absent = "penalize"))

sort(fit_big6$abilities, decreasing = TRUE)

sig6 <- bt_significance_matrix(fit_big6)
sig6
```

Of the 15 pairs, six are significant after Holm correction: Tottenham
against each of the other five clubs (all $p < 0.02$), plus Manchester
City vs. Manchester United ($p = 0.0105$). The remaining nine pairs —
among Manchester United, Arsenal, Chelsea and Liverpool, and between
Manchester City and each of those three — are not significant.
Tottenham is reliably behind the group; Manchester City is reliably
behind Manchester United specifically, but statistically
indistinguishable from Arsenal, Chelsea or Liverpool. This is exactly
the kind of claim a ranking table alone cannot support, and the reason
`bt_significance_matrix()` exists as a separate step from `bt_rank()`.

## Visualizing model fit: `plot.btfit()`

`plot.btfit(x, level = 0.95, ...)` draws a caterpillar plot: one point
per item (ability), one horizontal segment per item (Wald interval),
ordered weakest-to-strongest; items in `separation_items` are drawn in
grey rather than black. It is a thin wrapper around `graphics::plot()`
using exactly the table `summary.btfit()` computes — no separate
calculation.

```{r plot-btfit, eval = FALSE}
plot(fit_points)
```

Rendered figures are not included in this vignette; see the accompanying
JSS manuscript for the visual output of this and the two matrix
functions above.

## External validation: comparison with Elo

This section is the one deliberate exception to using only
**btrank** functions: its purpose is an independent check computed
*outside* the package. Elo (Elo, 1978) shares the same logistic
win-probability form as Bradley-Terry, so feeding an Elo rating into
`btrank` as a `score_col` would compare a Bradley-Terry fit against a
transformation of another Bradley-Terry-equivalent fit — not an
independent check.

Elo requires match-level results in chronological order, which
`epl_scores` does not retain. The ratings below were computed once via
`data-raw/compute_epl_elo.R` (starting rating 1500, $K=20$, no home
advantage, sequential updates over 11,780 matches) and are embedded
here rather than recomputed live.

```{r elo-comparison}
elo_final <- c(
  "Man City" = 1774.4, "Arsenal" = 1794.8, "Liverpool" = 1685.2,
  "Man United" = 1681.5, "Chelsea" = 1611.7, "Tottenham" = 1522.9,
  "Aston Villa" = 1667.8, "Newcastle" = 1599.4, "Brentford" = 1606.5,
  "Brighton" = 1615.0, "Bournemouth" = 1648.1, "Everton" = 1575.4,
  "West Ham" = 1538.9, "Crystal Palace" = 1571.8, "Fulham" = 1587.1,
  "Sunderland" = 1549.6, "Nott'm Forest" = 1580.6, "Wolves" = 1457.4,
  "Leeds" = 1554.0, "Burnley" = 1411.3
)

cor(rank(-fit_recent$abilities)[names(elo_final)], rank(-elo_final), method = "spearman")
```

Spearman correlation $\approx 0.64$ among the 20 current teams —
markedly lower than under `absent = "ignore"` weighting, since
`"penalize"` and the sequential, form-sensitive nature of Elo respond
very differently to a team's absence. Agreement at the very top still
holds (Manchester City and Arsenal occupy the top two positions under
both methods); the disagreement is concentrated further down the table.
Offered as a limitation worth noting, not resolved in favour of either
method.

## Relationship to BradleyTerry2

The second deliberate exception: demonstrating that `bt_fit()` is a
faithful wrapper, not an alternative estimator, necessarily calls
`BradleyTerry2::BTm()` directly.

```{r vs-bradleyterry2}
items <- rownames(win_mat_points)
pairs <- combn(items, 2)
df_manual <- data.frame(
  item1 = factor(pairs[1, ], levels = items), item2 = factor(pairs[2, ], levels = items),
  wins1 = apply(pairs, 2, function(p) win_mat_points[p[1], p[2]]),
  wins2 = apply(pairs, 2, function(p) win_mat_points[p[2], p[1]])
)
fit_manual <- BradleyTerry2::BTm(outcome = cbind(wins1, wins2), player1 = item1, player2 = item2, data = df_manual)

ability_manual <- BradleyTerry2::BTabilities(fit_manual)[, 1]
ability_btrank <- fit_points$abilities[names(ability_manual)]
common_ref <- names(ability_manual)[1]
max(abs((ability_manual - ability_manual[common_ref]) - (ability_btrank - ability_btrank[common_ref])))
```

At floating-point noise, once both fits share a reference (`BTm()`
fixes the first factor level; `bt_fit()` re-references to the weakest
item — comparing raw abilities without accounting for this produces a
constant, non-zero offset, not disagreement). What `btrank` adds on top
of `BradleyTerry2` is the layer documented above: deriving comparisons
from repeated scores (`bt_win_matrix()`), temporal decay
(`bt_weights()`), structural absence handling (`absent`), and
separation-aware uncertainty (`summary.btfit()`,
`bt_significance_matrix()`) — none of which `BradleyTerry2` provides.

## Limitations

- **Standard errors are not overdispersion-corrected.** Comparisons
  are round-robin-derived from a single season-aggregate score, not
  independent Bernoulli trials in the strict sense; SEs should be read
  as approximate.
- **`half_life` is a modelling choice**, not estimated — results should
  be checked across a range, not reported from one value.
- **No home advantage term.**
- **Separation is flagged, not corrected** — no bias-reduced estimator
  (e.g. Firth's correction) is applied.
- **Season-level aggregation discards match-level structure** — two
  teams finishing level on points are tied regardless of how they got
  there.

## References

Bradley, R. A., & Terry, M. E. (1952). Rank analysis of incomplete block
designs: I. The method of paired comparisons. *Biometrika*, 39(3/4), 324–345.

Turner, H., & Firth, D. (2012). Bradley-Terry models in R: The BradleyTerry2
package. *Journal of Statistical Software*, 48(9), 1–21.
https://doi.org/10.18637/jss.v048.i09

Ley, C., Van de Wiele, T., & Van Eetvelde, H. (2019). Ranking soccer teams
on the basis of their current strength: A comparison of maximum likelihood
approaches. *Statistical Modelling*, 19(1), 55–77.

Holm, S. (1979). A simple sequentially rejective multiple test procedure.
*Scandinavian Journal of Statistics*, 6(2), 65–70.

Liu, H. H., Battaglia, J., & Wu, T. T. (2025). Impact of COVID-19 on
English Football Premier League: Analyzing rankings and home advantage
using extended Bradley-Terry models. *PLOS One*, 20(10), e0332627.
https://doi.org/10.1371/journal.pone.0332627

Firth, D. (1993). Bias reduction of maximum likelihood estimates.
*Biometrika*, 80(1), 27–38.

Elo, A. E. (1978). *The Rating of Chessplayers, Past and Present*. Arco
Publishing, New York.

football-data.co.uk. English Premier League results, 1995/96–2025/26.
https://www.football-data.co.uk/englandm.php