---
title: "btrank: Bradley-Terry Rankings for Paired Comparison Data"
author: "Álvaro Meca Mondéjar, Miguel Ángel Montero Alonso, Juan de Dios Luna del Castillo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{btrank: Bradley-Terry Rankings for Paired Comparison Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE,
  message  = FALSE
)
```

## Overview

The **btrank** package provides a complete pipeline for estimating latent
ability scores from paired comparison data using the **Bradley-Terry model**
(Bradley & Terry, 1952). In a Bradley-Terry model, each item (team, player,
product, candidate) is assigned a latent ability parameter $\lambda_i$.
Given two items $i$ and $j$, the probability that $i$ outperforms $j$ is:

$$P(i \text{ beats } j) = \frac{e^{\lambda_i}}{e^{\lambda_i} + e^{\lambda_j}}$$

The model is estimated from pairwise comparisons aggregated across multiple
periods (seasons, years, rounds), making it particularly suitable for
ranking items whose relative strength evolves over time. **btrank** extends
the standard framework with:

- Support for **multiple metrics** as the basis for pairwise comparisons.
- **Temporal weighting** via exponential decay (Ley et al., 2019), allowing
  recent periods to carry more influence. Each period $t$ receives a weight

$$w_t = \left(\frac{1}{2}\right)^{(T - t) \,/\, T_{1/2}}$$

  where $T$ is the most recent period and $T_{1/2}$ is the half-life
  parameter controlling the rate of decay. The most recent period always
  has weight 1.
- Explicit handling of items absent from a given period (`absent`
  argument), relevant whenever the item universe changes across periods —
  promotion and relegation being a natural example, demonstrated below.
- A clean, pipe-friendly API separating data ingestion, model fitting and
  result extraction into modular functions.

This vignette demonstrates the full pipeline on a real dataset: 31 seasons
of English Premier League results (1995/96–2025/26), chosen because its
outcomes are independently verifiable against public record.

## Installation

```{r install, eval = FALSE}
remotes::install_github("alvaromkm/btrank")
```

```{r load}
library(btrank)
```

## Input data format

**btrank** expects a data frame with one row per item per period. The
minimum required columns are:

| Column | Description |
|--------|-------------|
| `item` | Item identifier (team, player, product, ...) |
| `period` | Period identifier (season, year, round, ...) |
| `score` | Numeric metric used to rank items within each period |

Column names are fully customisable via the `item_col`, `period_col` and
`score_col` arguments. By default, higher values of `score` indicate better
performance. Set `higher_is_better = FALSE` for metrics where lower is
better (e.g. time, errors).

## The EPL dataset

```{r load-data}
dim(epl_scores)
head(epl_scores, 8)
```

Each row is one team's aggregate record for one season: `points` (3/1/0 per
win/draw/loss) and `goal_diff` (goals for minus goals against), both
already in the `item`/`period`/`score` format `bt_win_matrix()` expects.

## Data provenance and reproducibility

Match results are sourced from
[football-data.co.uk](https://www.football-data.co.uk/englandm.php)
(division "E0"). Raw match-level CSV files are **not** redistributed with
this package, as their exact redistribution terms are not stated by the
source; only the aggregated `epl_scores` table is shipped as package data.
The aggregation itself — from raw match CSVs to `item`/`period`/`points`/
`goal_diff` — is implemented in the internal, unit-tested function
`.matches_to_scores()`, and the exact steps to regenerate `epl_scores` from
the original source are documented in `data-raw/prepare_epl_scores.R`,
included in the package source repository.

## Fitting the model: league points

Following the same two-step workflow shown above:

```{r fit-points}
win_mat_points <- bt_win_matrix(epl_scores, score_col = "points")
fit_points      <- bt_fit(win_mat_points)
print(fit_points)

bt_rank(fit_points)[1:10, ]
```

Across the full 31-season window, with no temporal weighting, the top of
the table is Manchester United, Arsenal, Liverpool, Chelsea, Manchester
City and Tottenham — the "big six" widely recognised in English football,
recovered here purely from 31 years of season-aggregate points, with no
football-specific knowledge built into the model.

The `ability_norm` column rescales the raw log-ability $\lambda_i$ to the
$[0, 1]$ interval, mapping the weakest item in the set to 0 and the
strongest to 1. Raw $\lambda_i$ values are not comparable across different
model fits (their scale depends on the number of items, periods, and
degree of separation between them); `ability_norm` is, making it the right
column to use when comparing rankings from different fits, as done below.

### Temporal weighting changes who leads

A flat, unweighted fit answers "who was strongest averaged over 31 years?".
That is a different question from "who is strongest *now*?". Compare a flat
fit against one with `half_life = 5`:

```{r weighted-fit}
periods    <- sort(unique(epl_scores$period))
w_recent   <- bt_weights(periods, half_life = 5)

win_mat_recent <- bt_win_matrix(epl_scores, score_col = "points", weights = w_recent)
fit_recent      <- bt_fit(win_mat_recent)

rk_flat   <- bt_rank(fit_points)
rk_recent <- bt_rank(fit_recent)

merge(
  data.frame(item = rk_flat$item,   rank_flat   = rk_flat$rank),
  data.frame(item = rk_recent$item, rank_recent = rk_recent$rank),
  by = "item"
)
```

Under `half_life = 5`, Manchester City moves from 5th to **1st**, and
Manchester United drops from 1st to 4th — reflecting Manchester City's
2017–2023 run of dominance against Manchester United's trophy-heavy but
now more distant 1995–2013 record. This is precisely the effect
[bt_weights()] is designed to capture, on data where the underlying
football history is independently checkable.

## Fitting the model: goal difference

Points is one lens on team strength; goal difference is another — it
rewards margin of victory rather than just the win/draw/loss outcome. The
same pipeline applies, only the `score_col` changes:

```{r fit-goal-diff}
win_mat_gd <- bt_win_matrix(epl_scores, score_col = "goal_diff")
fit_gd      <- bt_fit(win_mat_gd)

bt_rank(fit_gd)[1:10, ]
```

### Where the two metrics agree, and where they don't

```{r compare-metrics}
rk_points <- bt_rank(fit_points)
rk_gd     <- bt_rank(fit_gd)

cmp <- merge(
  data.frame(item = rk_points$item, rank_points    = rk_points$rank),
  data.frame(item = rk_gd$item,     rank_goal_diff = rk_gd$rank),
  by = "item"
)
cmp <- cmp[order(cmp$rank_points), ]

cor(cmp$rank_points, cmp$rank_goal_diff, method = "spearman")

head(cmp, 15)
```

The two rankings agree closely overall (Spearman correlation ≈ 0.97) — the
top six are the same set of clubs under either metric. But the *order*
within that top six is not identical: Arsenal ranks 1st by goal difference
but 2nd by points, while Manchester United is the reverse. Further down,
Brentford ranks 7th by goal difference but only 11th by points — a team
whose results were closer than their points tally alone suggests. Teams
such as Charlton show the opposite pattern: a points total propped up by
close, low-margin results loses ground once judged by goal difference
(19th by points, 30th by goal difference). This is the intended effect of
supporting multiple metrics: the two questions ("who won more?" and "who
dominated more?") are related but not identical, and **btrank** lets the
user ask either one from the same `item`/`period`/score` input.

## Handling absence: promotion and relegation

The full 31-season item universe includes teams that were only ever in the
Premier League decades apart from each other. As `bt_win_matrix()`'s
documentation notes, `absent` and the time window are linked: passing an
unbounded window with `absent = "penalize"` accumulates structural losses
for spans of absence that were never the intended comparison. This section
therefore bounds the window to the five most recent seasons (2021–2025),
where promotion and relegation are directly observable.

```{r absent-setup}
recent <- epl_scores[epl_scores$period %in% 2021:2025, ]

# Teams present in fewer than all 5 of these seasons
tab <- table(recent$item, recent$period)
partial <- rowSums(tab > 0)
partial[partial < 5]
```

### `absent = "ignore"` (default)

```{r absent-ignore, warning = TRUE}
win_mat_ignore <- bt_win_matrix(recent, score_col = "points", absent = "ignore")
fit_ignore      <- bt_fit(win_mat_ignore)

bt_rank(fit_ignore)[1:6, ]
```

Fitting raises a separation warning for Norwich and Sheffield United — each
played only one of these five seasons and lost every within-season
comparison, so their ability estimate is a boundary point estimate with an
unreliable standard error (see `?bt_fit`, Details).

More subtly, notice **Sunderland ranks 4th**, just behind Liverpool and
ahead of Manchester United — despite having played only one of these five
seasons (2024/25, their first back in the Premier League after several
years in the Championship). Under `"ignore"`, an item's ability reflects
only the periods it was observed in; the four seasons Sunderland was absent
contribute nothing, positive or negative, to its estimate.

### `absent = "penalize"`

```{r absent-penalize}
win_mat_penalize <- bt_win_matrix(recent, score_col = "points",
                                   absent = "penalize", absent_penalty = 1)
fit_penalize      <- bt_fit(win_mat_penalize)

rank(-fit_penalize$abilities)["Sunderland"]
fit_penalize$abilities["Sunderland"]
```

With every team present each season credited a structural win against
Sunderland for the four seasons it was absent, Sunderland's ability drops
sharply and its rank falls from 4th to **20th** — a considerably more
defensible position given its actual top-flight record over this window.

### Choosing between them

Neither setting is a free default: `"ignore"` can let a single strong
season outweigh years of absence, as above; `"penalize"` treats absence
itself as informative, which is a modelling assumption that should be
stated, not left implicit (see `?bt_win_matrix`, `absent_penalty`). An
alternative used elsewhere in the literature is to exclude teams without
consistent presence across the window entirely (Liu, Battaglia & Wu, 2025)
— a stricter but simpler choice than penalizing.

## External validation: comparison with Elo

Elo (Elo, 1978) is a widely used alternative rating system for paired
comparisons, updated sequentially match by match rather than fit in a
single batch. Because its underlying win-probability model is the same
logistic form as Bradley-Terry, feeding an Elo rating into `btrank` as a
`score_col` would compare a Bradley-Terry fit against a transformation of
another Bradley-Terry-equivalent fit — not an independent check. Elo is
therefore used here purely as an **external** point of comparison, computed
entirely outside the `btrank` pipeline.

Computing Elo requires match-level results in chronological order, which
`epl_scores` does not retain (only season aggregates are shipped — see
"Data provenance" above). The ratings below were computed once, with a
standard implementation (starting rating 1500, $K = 20$, no home-advantage
adjustment, sequential updates over all 11,780 matches from 1995/96 to
2025/26 in date order); the script is included as
`data-raw/compute_epl_elo.R` for reproducibility, and its output is
embedded below rather than recomputed live in this vignette.

```{r elo-comparison}
# Final Elo rating (end of 2025/26), current 20 Premier League teams.
# Computed via data-raw/compute_epl_elo.R; see that script for the full,
# match-level reproduction from the original source.
elo_final <- c(
  "Man City" = 1774.4, "Arsenal" = 1794.8, "Liverpool" = 1685.2,
  "Man United" = 1681.5, "Chelsea" = 1611.7, "Tottenham" = 1522.9,
  "Aston Villa" = 1667.8, "Newcastle" = 1599.4, "Brentford" = 1606.5,
  "Brighton" = 1615.0, "Bournemouth" = 1648.1, "Everton" = 1575.4,
  "West Ham" = 1538.9, "Crystal Palace" = 1571.8, "Fulham" = 1587.1,
  "Sunderland" = 1549.6, "Nott'm Forest" = 1580.6, "Wolves" = 1457.4,
  "Leeds" = 1554.0, "Burnley" = 1411.3
)

rk_recent_named <- setNames(rk_recent$rank, rk_recent$item)

cmp_elo <- data.frame(
  item        = names(elo_final),
  rank_btrank = rk_recent_named[names(elo_final)],
  rank_elo    = rank(-elo_final)
)
cmp_elo <- cmp_elo[order(cmp_elo$rank_btrank), ]

cor(cmp_elo$rank_btrank, cmp_elo$rank_elo, method = "spearman")

cmp_elo
```

The two independently-derived rankings agree closely (Spearman correlation
≈ 0.80) among the 20 teams currently in the league, and agree almost
exactly at the top: Manchester City and Arsenal occupy the top two
positions under both methods (swapped), as do Liverpool and Manchester
United in 3rd–4th. The clearest disagreement is Tottenham — 6th under
`btrank`'s decayed points ranking but only 18th by Elo — a genuine
divergence between a method built on season-aggregate results and one
built on sequential match-by-match form, offered here as a limitation
worth noting rather than resolved in favour of either.


## Relationship to BradleyTerry2

**btrank** does not reimplement Bradley-Terry estimation: every model fit
in this vignette is, underneath, a call to `BradleyTerry2::BTm()` (Turner &
Firth, 2012). This section makes that explicit, and clarifies what
`bt_win_matrix()` and `bt_fit()` add on top of it.

### Same estimator, same numbers

Given an already-built win count matrix, fitting "by hand" with
`BradleyTerry2::BTm()` directly reproduces `bt_fit()`'s point estimates
exactly:

```{r vs-bradleyterry2}
# Manual replication of what bt_fit() does internally, using BradleyTerry2
# directly, with no help from btrank beyond the win matrix already built.
items <- rownames(win_mat_points)
pairs <- combn(items, 2)
df_manual <- data.frame(
  item1 = factor(pairs[1, ], levels = items),
  item2 = factor(pairs[2, ], levels = items),
  wins1 = apply(pairs, 2, function(p) win_mat_points[p[1], p[2]]),
  wins2 = apply(pairs, 2, function(p) win_mat_points[p[2], p[1]])
)

fit_manual <- BradleyTerry2::BTm(
  outcome = cbind(wins1, wins2), player1 = item1, player2 = item2,
  data = df_manual
)

ability_manual <- BradleyTerry2::BTabilities(fit_manual)[, 1]
ability_btrank <- fit_points$abilities[names(ability_manual)]

# BT abilities are identified only up to an additive constant: BTm() and
# bt_fit() use different reference items by default (BTm() fixes the first
# factor level; bt_fit() re-references to the weakest item). Put both on
# the same baseline — any common item works — before comparing.
common_ref <- names(ability_manual)[1]
max(abs(
  (ability_manual - ability_manual[common_ref]) -
  (ability_btrank - ability_btrank[common_ref])
))
```

The maximum difference, once both fits are placed on the same reference
item, is at floating-point noise level: `bt_fit()` is a thin, faithful
wrapper around `BTm()`, not an alternative estimator. (The two fits use
different default reference items — `BTm()` fixes the first factor level,
`bt_fit()` re-references to the weakest item — which is why comparing raw
`abilities` vectors directly, without accounting for this, produces a
constant, non-zero offset rather than agreement.)

### What `BradleyTerry2` does not provide

`BradleyTerry2::BTm()` takes a pairwise win count matrix as given — it has
no notion of *deriving* one from repeated, dated observations of a metric.
Everything upstream of `win_mat_points` in this vignette has no equivalent
in `BradleyTerry2`:

- **Turning season-level scores into comparisons.** `bt_win_matrix()`
  converts one row per item per period into the pairwise structure `BTm()`
  needs. Without it, a user would derive round-robin comparisons from raw
  scores by hand for every dataset.
- **Temporal decay.** `bt_weights()` and the `weights` argument of
  `bt_win_matrix()` are what produced the Manchester City / Manchester
  United reordering shown earlier. `BradleyTerry2` has no period concept
  and no decay mechanism; every comparison is treated as equally
  informative regardless of when it occurred.
- **Structural handling of absence.** The `absent` argument — demonstrated
  above with Sunderland — has no counterpart in `BradleyTerry2`; a
  promoted or relegated item is simply missing rows, silently.
- **Separation-aware uncertainty quantification.** `summary.btfit()`,
  `confint.btfit()` and `bt_significance_matrix()` build on `BTm()`'s own
  `glm()` machinery, but add automatic detection of quasi-complete
  separation (as with Norwich and Sheffield United above) and flag which
  standard errors and pairwise tests should not be trusted as a result —
  a check `BradleyTerry2` leaves entirely to the user.

**btrank**'s contribution is this layer: a data pipeline and safety checks
built around `BradleyTerry2`'s estimator, not a replacement for it.


## Limitations

This vignette demonstrates the pipeline on real data with results that are
independently checkable, but several modelling choices made along the way
are worth stating explicitly rather than leaving implicit.

- **Standard errors are not overdispersion-corrected.** `bt_fit()`'s
  uncertainty quantification relies on `BTm()`'s underlying `glm()`
  machinery, which assumes each recorded comparison is independent. The
  comparisons `bt_win_matrix()` generates are round-robin-derived from a
  single season-aggregate score per item per period, not independent
  observations in the strict sense — closer in spirit to Turner & Firth's
  (2012, Section 9.1) tie-splitting device than to genuinely independent
  Bernoulli trials. Standard errors reported by `summary.btfit()` and
  `confint.btfit()` should therefore be read as approximate, and are most
  useful as a relative guide (which differences are large versus small)
  rather than as calibrated interval estimates.

- **`half_life` is a modelling choice, not an estimated quantity.**
  `half_life = 5` was chosen above to illustrate the effect of temporal
  weighting; nothing in the data selects that value over `half_life = 3`
  or `half_life = 8`. As with `absent_penalty` (see `?bt_win_matrix`),
  results should be checked across a range of values rather than reported
  from a single choice.

- **No home advantage term.** Both `points` and `goal_diff` are
  season-level aggregates that already mix home and away results, but
  neither `bt_win_matrix()` nor `bt_fit()` models home advantage
  explicitly as a separate effect, unlike extended Bradley-Terry
  approaches built for this purpose (e.g. Liu, Battaglia & Wu, 2025).

- **Quasi-complete separation is flagged, not resolved.** As shown with
  Norwich and Sheffield United, `bt_fit()` reports which items have
  unreliable standard errors; it does not apply a penalised or
  bias-reduced estimator (e.g. Firth's correction) to fix the underlying
  problem. The point estimate for a separated item is still usable, its
  associated uncertainty is not.

- **Season-level aggregation discards match-level structure.** Ranking is
  based on one score per item per period, not on the individual matches
  behind it. This is a deliberate trade-off — it is what makes temporal
  decay across periods well-defined — but it means, for instance, that two
  teams finishing a season with identical points are treated as tied
  regardless of how they arrived there.

## References

Bradley, R. A., & Terry, M. E. (1952). Rank analysis of incomplete block
designs: I. The method of paired comparisons. *Biometrika*, 39(3/4), 324–345.

Turner, H., & Firth, D. (2012). Bradley-Terry models in R: The BradleyTerry2
package. *Journal of Statistical Software*, 48(9), 1–21.
https://doi.org/10.18637/jss.v048.i09

Ley, C., Van de Wiele, T., & Van Eetvelde, H. (2019).
Ranking soccer teams on the basis of their current strength: A comparison
of maximum likelihood approaches. *Statistical Modelling*, 19(1), 55–73.

Liu, H. H., Battaglia, J., & Wu, T. T. (2025). Impact of COVID-19 on
English Football Premier League: Analyzing rankings and home advantage
using extended Bradley-Terry models. *PLOS One*, 20(10), e0332627.
https://doi.org/10.1371/journal.pone.0332627

Elo, A. E. (1978). *The Rating of Chessplayers, Past and Present*. Arco
Publishing, New York.

football-data.co.uk. English Premier League results, 1995/96–2025/26.
https://www.football-data.co.uk/englandm.php