---
title: "Introduction to btrank: Bradley-Terry Rankings for Paired Comparison Data"
author: "Álvaro Meca Mondéjar, Miguel Ángel Montero Alonso, Juan de Dios Luna del Castillo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to btrank: Bradley-Terry Rankings for Paired Comparison Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE,
  message  = FALSE
)
```

## Overview

The **btrank** package provides a complete pipeline for estimating latent 
ability scores from paired comparison data using the **Bradley-Terry model** 
(Bradley & Terry, 1952). 


In a Bradley-Terry model, each item (team, player, product, candidate) is 
assigned a latent ability parameter $\lambda_i$. Given two items $i$ and $j$, 
the probability that $i$ outperforms $j$ is:

$$P(i \text{ beats } j) = \frac{e^{\lambda_i}}{e^{\lambda_i} + e^{\lambda_j}}$$


The model is estimated from pairwise comparisons aggregated across multiple 
periods (seasons, years, rounds). This makes it particularly suitable for 
ranking items whose relative strength evolves over time.


**btrank** extends the standard Bradley-Terry framework with:


- Support for **multiple metrics** as the basis for pairwise comparisons.


- **Temporal weighting** via exponential decay (Ley et al., 2019), allowing 
  recent periods to carry more influence. Each period $t$ receives a weight:

$$w_t = \left(\frac{1}{2}\right)^{(T - t) \,/\, T_{1/2}}$$

where $T$ is the most recent period and $T_{1/2}$ is the half-life parameter  controlling the rate of decay. The most recent period always has weight 1.
  
  
- A clean, pipe-friendly API that separates data ingestion, model fitting and result extraction into modular functions.
  

## Installation

```{r install, eval = FALSE}
# Install from GitHub
remotes::install_github("alvaromkm/btrank")
```

```{r load}
library(btrank)
```


## Input data format

**btrank** expects a data frame with one row per item per period. The minimum 
required columns are:


| Column | Description |
|--------|-------------|
| `item` | Item identifier (team, player, product, ...) |
| `period` | Period identifier (season, year, round, ...) |
| `score` | Numeric metric used to rank items within each period |


Column names are fully customisable via the `item_col`, `period_col` and 
`score_col` arguments. By default, higher values of `score` indicate better 
performance. Set `higher_is_better = FALSE` for metrics where lower is better 
(e.g. time, errors).


## Basic pipeline

The package provides two workflows depending on the level of control needed.

### One-step workflow

For most use cases, `bt_rank_all()` runs the full pipeline in a single call:

```{r basic-example}
# Five chess players across four tournaments
chess <- data.frame(
  item   = c("Carlsen", "Caruana", "Ding", "Nepomniachtchi", "Gukesh",
             "Carlsen", "Caruana", "Ding", "Nepomniachtchi", "Gukesh",
             "Carlsen", "Caruana", "Ding", "Nepomniachtchi", "Gukesh",
             "Carlsen", "Caruana", "Ding", "Nepomniachtchi", "Gukesh"),
  period = c(rep(2020, 5), rep(2021, 5), rep(2022, 5), rep(2023, 5)),
  score  = c(87, 82, 74, 78, 65,
             89, 80, 76, 75, 68,
             91, 83, 77, 73, 72,
             90, 84, 75, 71, 78)
)

bt_rank_all(chess, score_col = "score")
```

### Step-by-step workflow

For more control, the pipeline can be run function by function:

```{r stepbystep}
# Step 1: build the pairwise win count matrix
win_mat <- bt_win_matrix(chess, score_col = "score")

# Step 2: fit the Bradley-Terry model
fit <- bt_fit(win_mat)
print(fit)

# Step 3: extract the ranking
bt_rank(fit)

# Step 4 (optional): compute win probability matrix
round(bt_prob_matrix(fit), 3)
```



## Temporal weighting

By default, all periods contribute equally to the model. When performance 
evolves over time — for instance, when younger competitors improve 
consistently across seasons — it may be desirable to assign greater influence 
to more recent periods.


`bt_weights()` computes exponential decay weights for a set of periods. The 
`half_life` parameter controls how quickly the influence of past periods fades:
a shorter half-life concentrates the ranking on recent performance, while a 
longer one treats all periods more uniformly.

```{r weights}
# Inspect weights for different half-life values
w_aggressive <- bt_weights(periods = 2020:2023, half_life = 1)
w_smooth     <- bt_weights(periods = 2020:2023, half_life = 3)

round(rbind(
  "half_life = 1" = w_aggressive,
  "half_life = 3" = w_smooth
), 4)
```

The weights can be passed directly to `bt_win_matrix()`, or implicitly via 
the `half_life` argument of `bt_rank_all()`:

```{r weighted-ranking}
# Without temporal weighting
rk_flat <- bt_rank_all(chess, score_col = "score")

# With temporal weighting — recent seasons matter more
rk_weighted <- bt_rank_all(chess, score_col = "score", half_life = 1)

# Compare rankings side by side
merge(
  data.frame(item = rk_flat$item,     rank_flat     = rk_flat$rank),
  data.frame(item = rk_weighted$item, rank_weighted = rk_weighted$rank),
  by = "item"
)
```

Notice how items whose performance improves in later periods gain positions 
when temporal weighting is applied. In this dataset, Gukesh scores 
consistently higher in 2022 and 2023 than in earlier periods, which is 
reflected in a better ranking under aggressive weighting.



## Win probability matrix

Once the model is fitted, `bt_prob_matrix()` computes the probability that 
any item beats any other item in a direct comparison. This provides a more 
nuanced view of competitive strength than a simple ranking — two items may be 
adjacent in the ranking but have very different win probabilities against 
specific opponents.

```{r prob-matrix}
win_mat <- bt_win_matrix(chess, score_col = "score")
fit     <- bt_fit(win_mat)
pm      <- bt_prob_matrix(fit)

round(pm, 3)
```

Entry `[i, j]` is the probability that item in row $i$ beats item in column 
$j$. The diagonal is `NA` by default. By construction, 
$P(i \text{ beats } j) + P(j \text{ beats } i) = 1$ for all pairs.


A useful summary is to identify dominant pairs — those where one item has a 
win probability above a chosen threshold:

```{r dominance}
threshold <- 0.75

dominant <- which(pm > threshold, arr.ind = TRUE)
if (nrow(dominant) > 0) {
  data.frame(
    winner   = rownames(pm)[dominant[, 1]],
    loser    = colnames(pm)[dominant[, 2]],
    prob_win = round(pm[dominant], 3)
  )
} else {
  message("No pairs exceed the threshold of ", threshold)
}
```

The matrix is ordered by estimated ability (strongest item first) by default. 
Use `order_by = "name"` for alphabetical ordering:

```{r prob-matrix-alpha}
round(bt_prob_matrix(fit, order_by = "name"), 3)
```


## Interpreting the ability parameter


The ability estimates $\lambda_i$ returned by `bt_rank()` are log-scale 
parameters relative to the weakest item in the set, which is always fixed 
at $\lambda = 0$. This means:


- A value of $\lambda_i = 0$ indicates the reference item (worst performer).


- All other values are **positive** and express strength relative to the 
  reference.
  
  
- The absolute magnitude of $\lambda_i$ depends on the number of items, 
  periods, and degree of separation between them — it is not directly 
  comparable across different model fits.
  

For cross-model comparisons, use the `ability_norm` column returned by 
`bt_rank()`, which rescales abilities to the $[0, 1]$ interval:

```{r ability-norm}
bt_rank(fit)
```

The `ability_norm` column maps the weakest item to 0 and the strongest to 1, 
making relative comparisons between different rankings more interpretable.


## References

Bradley, R. A., & Terry, M. E. (1952). Rank analysis of incomplete block 
designs: I. The method of paired comparisons. *Biometrika*, 39(3/4), 324–345.


Ley, C., Van Eetvelde, M., Claeskens, G., & Winkler, R. L. (2019). Ranking 
soccer teams on the basis of their current strength: A comparison of maximum 
likelihood approaches. *Statistical Modelling*, 19(1), 55–73.


Alvo, M., & Yu, P. L. H. (2014). *Statistical Methods for Ranking Data*. 
Springer, New York. https://doi.org/10.1007/978-1-4939-1471-5


Yu, P. L. H., Gu, J., & Xu, H. (2019). Analysis of ranking data. 
*WIREs Computational Statistics*. https://doi.org/10.1002/wics.1483
